---
title: "Project"
author: "Theo and Thomas"
date: "October 2025"
output:
  pdf_document:
    toc: false
    number_sections: false
fontsize: 11pt
geometry: margin=1in
urlcolor: blue
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6, fig.height = 4)
library(tidyverse)
library(knitr)
library(readxl)
library(MASS)
library(dplyr)
library(car)
library(broom)
library(GGally)
library(glmmTMB)
library(logistf)
library(ggplot2)
library(ResourceSelection)
```

Interval world_data 
```{r}
interval_data = read.csv("FINAL DATA.csv")

interval_data   
```

Create world_data 
```{r}
#Read in world_data  (world_data  is ordered)
world_data   = read.csv("FINAL WORLD DATA.csv")

# Create composite features
world_data  = world_data  %>%
  mutate(
    # Overall success rate up to W8_3
    success_rate = (W1_1 + W1_2 + W4_1 + W4_2 + W8_1 + W8_2 + W8_3) / 7
  )

world_data 
```

Create time varied world_data 
```{r}
# Create day–level improvement variable

day_summary = world_data  %>%
  group_by(Day) %>%
  summarise(
    day_success_rate = mean(success_rate),
    .groups = "drop"
  ) %>%

  mutate(
    time_index = as.numeric(factor(Day))
  ) %>%
  arrange(time_index) %>%

  mutate(
    day_improvement = day_success_rate - lag(day_success_rate),
    day_cum_improvement = day_success_rate - first(day_success_rate)
  )


world_data  = world_data  %>%
  left_join(day_summary, by = "Day")

world_data $day_improvement[is.na(world_data $day_improvement)] = 0

world_data 

#Graph day_success_rate over time grouped by date to find IF there is improvement over time


```
Create model for improvement over time
```{r}
interval_data = interval_data   %>%
  mutate(DayNum = as.numeric(gsub('Day_', '', Day)))

world_data = world_data  %>%
  mutate(DayNum = as.numeric(gsub('Day_', '', Day)))

# Calculate success rate by day and interval
daily_interval_success = interval_data   %>%
  group_by(DayNum, Interval) %>%
  summarise(
    success_rate = mean(Success),
    n_attempts = n(),
    .groups = 'drop'
  )

# Calculate overall success rate by day (across all intervals)
daily_overall_success = interval_data   %>%
  group_by(DayNum) %>%
  summarise(
    overall_success_rate = mean(Success),
    total_attempts = n(),
    .groups = 'drop'
  )

# Calculate how far Nifski get on average each day
daily_progression = interval_data   %>%
  group_by(DayNum, Run) %>%
  summarise(
    max_interval_reached = max(Interval[Success == 1]),
    .groups = 'drop'
  ) %>%
  group_by(DayNum) %>%
  summarise(
    avg_max_interval = mean(max_interval_reached),
    median_max_interval = median(max_interval_reached),
    .groups = 'drop'
  )
```

Graphs
```{r}
# Graph 1: Overall Success Rate Over Time (Enhanced with Legend)
ggplot(daily_overall_success, aes(x = DayNum, y = overall_success_rate)) +
  geom_point(aes(color = "Daily Success Rate"), size = 4, alpha = 0.7) +
  geom_smooth(aes(color = "Smoothed Trend"), 
              method = "loess", se = FALSE, linewidth = 1.5) +
  geom_smooth(aes(color = "Linear Trend"), 
              method = "lm", se = TRUE, linewidth = 1.2, alpha = 0.2) +
  scale_color_manual(name = "Legend",
                     values = c("Daily Success Rate" = "steelblue",
                               "Smoothed Trend" = "darkblue",
                               "Linear Trend" = "red")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), 
                     limits = c(0.50, 0.70)) +
  scale_x_continuous(breaks = seq(0, 21, by = 3)) +
  labs(title = "Speedrun Performance Improvement Over Time",
       subtitle = "Overall success rate across all intervals by day",
       x = "Day",
       y = "Success Rate") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray30"),
        panel.grid.minor = element_blank(),
        axis.title = element_text(face = "bold"),
        legend.position = "right")
```

Models
```{r}
#Now that we've shown that there is improvement
# Model improvement over time on interval completion
cloglog_time = glm(Success ~ DayNum + factor(Interval), 
                   data = interval_data, 
                   family = binomial(link = "cloglog"))
summary(cloglog_time)

cloglog_full = glm(Success ~ DayNum * factor(Interval), 
                   data = interval_data, 
                   family = binomial(link = "cloglog"))

anova(cloglog_time, cloglog_full, test = "Chisq")
#saturated model is better so do stepwise to find best AIC

```

Stepwise
```{r}
# Null model (intercept only)
cloglog_null = glm(Success ~ 1, 
                   data = interval_data, 
                   family = binomial(link = "cloglog"))

# Forward stepwise selection
forward_model = step(cloglog_null, 
                     scope = list(lower = cloglog_null, upper = cloglog_full),
                     direction = "forward",
                     trace = 0)

# Backward stepwise selection
backward_model = step(cloglog_full,
                      direction = "backward",
                      trace = 0)

# Both directions stepwise selection
both_model = step(cloglog_null,
                  scope = list(lower = cloglog_null, upper = cloglog_full),
                  direction = "both",
                  trace = 0)

best_model = forward_model

summary(best_model)
#Best model is Success ~ factor(Interval) + DayNum + factor(Inveral):DayNum

```

Find if model is a good fit
```{r}
# Analyze best_model fit

cat("P-value:", deviance_gof, "\n")
cat("Interpretation: p >0.05 indicates good fit\n")

# 3. Pearson Chi-Square Test
pearson_resid = residuals(best_model, type = "pearson")
pearson_chisq = sum(pearson_resid^2)
pearson_pval = 1 - pchisq(pearson_chisq, best_model$df.residual)
cat("\nPearson Chi-Square Test:\n")
cat("Chi-Square:", pearson_chisq, "\n")
cat("P-value:", pearson_pval, "\n")

# 4. Hosmer-Lemeshow Test
library(ResourceSelection)
hl_test = hoslem.test(interval_data$Success, fitted(best_model), g = 10)
print(hl_test)
cat("Hosmer-Lemeshow: p >0.05 indicates good fit\n")

# 5. McFadden's Pseudo R-squared
null_dev = best_model$null.deviance
resid_dev = best_model$deviance
pseudo_r2 = 1 - (resid_dev / null_dev)
cat("\nMcFadden's Pseudo R-squared:", pseudo_r2, "\n")
cat("Interpretation: 0.2-0.4 = excellent fit\n")

# 6. Calculate VIF for multicollinearity (if applicable)
cat("\nVariance Inflation Factors:\n")
vif_values = vif(best_model)
print(vif_values)
cat("VIF < 10 indicates no serious multicollinearity\n")

# 7. GRAPHS

# Graph 3: Residuals by Interval
ggplot(interval_data, aes(x = factor(Interval), y = deviance_resid)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals by Interval",
       subtitle = "Checking if residuals vary by interval level",
       x = "Interval",
       y = "Deviance Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
#This shows that the interaction term is fine, Interval 3,4,7 might have more to it that my model doesn't capture


# Graph 4: Residuals over Time
ggplot(interval_data, aes(x = DayNum, y = deviance_resid)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "darkblue") +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals Over Time",
       subtitle = "Checking for temporal patterns",
       x = "Day",
       y = "Deviance Residuals") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# 8. Influential Points
interval_data$cooks_d = cooks.distance(best_model)

# Graph 5: Cook's Distance
ggplot(interval_data, aes(x = 1:nrow(interval_data), y = cooks_d)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = 4/nrow(interval_data), color = "red", 
             linetype = "dashed", linewidth = 1) +
  labs(title = "Cook's Distance - Influential Observations",
       subtitle = "Points above red line may be influential",
       x = "Observation Index",
       y = "Cook's Distance") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# 9. Summary Statistics
cat("\n=== MODEL FIT SUMMARY ===\n")
cat("AIC:", AIC(best_model), "\n")
cat("BIC:", BIC(best_model), "\n")
cat("Log-Likelihood:", logLik(best_model), "\n")
cat("Pseudo R-squared:", pseudo_r2, "\n")

```

Fitted vs Actual Graph
```{r}
# Better: Direct Actual vs Fitted comparison

# Create a dataset with both actual and fitted values
comparison_data = interval_data %>%
  dplyr::select(DayNum, Interval, Success, fitted_prob) %>%
  arrange(fitted_prob)

# Add observation index
comparison_data$obs_index = 1:nrow(comparison_data)

#Scatterplot with jitter (simpler, no select needed)
ggplot(interval_data, aes(x = fitted_prob, y = Success)) +
  geom_jitter(alpha = 0.3, width = 0, height = 0.05, color = "steelblue") +
  geom_smooth(method = "loess", se = TRUE, color = "red", linewidth = 1.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "darkgreen") +
  labs(title = "Actual vs Fitted Probabilities",
       subtitle = "Red line shows actual relationship, green line is perfect prediction",
       x = "Fitted Probability",
       y = "Actual Outcome (0 or 1)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))




# Option 2: ROC-style cumulative plot
interval_data_sorted = interval_data %>%
  arrange(fitted_prob) %>%
  mutate(
    obs_number = row_number(),
    cumulative_fitted = cummean(fitted_prob),
    cumulative_actual = cummean(Success)
  )

ggplot(interval_data_sorted, aes(x = obs_number)) +
  geom_line(aes(y = cumulative_actual, color = "Actual (Cumulative Avg)"), 
            linewidth = 1.2) +
  geom_line(aes(y = cumulative_fitted, color = "Fitted (Cumulative Avg)"), 
            linewidth = 1.2) +
  scale_color_manual(name = "Value",
                     values = c("Actual (Cumulative Avg)" = "red",
                               "Fitted (Cumulative Avg)" = "blue")) +
  labs(title = "Cumulative Average: Actual vs Fitted",
       subtitle = "Observations sorted by fitted probability",
       x = "Observation (sorted by fitted probability)",
       y = "Cumulative Average") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom")

```

Use Predicted Success Rate from Interval Model
```{r}
best_wr_model = glm(W8_4 ~ success_rate + DayNum,
                    data = world_data,
                    family = binomial(link = "cloglog"))
best_wr_model2 = glm(W8_4 ~  DayNum,
                    data = world_data,
                    family = binomial(link = "cloglog"))

summary(best_wr_model)
summary(best_wr_model2)


AIC(best_wr_model)


world_data$wr_predicted_prob = predict(best_wr_model, type = "response")

# Example: Predict for perfect run on Day 22
perfect_run = data.frame(
  success_rate = 1.0,
  DayNum = 22
)
predict(best_wr_model, newdata = perfect_run, type = "response")
anova(best_wr_model2, best_)

#SCRAP SUCCESS RATE
#Predict improvement rate

#To predict improvement rate build models where response is Interval 1, then Interval 2 etc... instead of predicting success
#Use improvement rate to build model for each level
#Then make deterministic probability based on these models

#Survivor model based on what we just went over in class, might be an issue that the survivors are really rare, might not be, have to test it

#Censored runs are the WR runs, we are trying to see if his lifetime goes up over time

```


ZIP
```{r}
#logistic binomial regression with ZIP 
zib_0 = glmmTMB(
  W8_4 ~ 1,        
  ziformula = ~ 1,
  family = binomial(link = logit),
  data  = world_data 
)

summary(zib_0)

zib_full = glmmTMB(
  W8_4 ~ W1_1 + W1_2 + W4_1 + W4_2 + W8_1 + W8_2 + W8_3,        
  ziformula = ~ W1_1 + W1_2 + W4_1 + W4_2 + W8_1 + W8_2 + W8_3, 
  family = binomial(link = logit),
  data  = world_data 
)
summary(zib_full)
#no warning but still NAN p values, ZIP doesn't work

zib_full_2 = glmmTMB(
  W8_4 ~ success_rate,        
  ziformula = ~ success_rate, 
  family = binomial(link = logit),
  data  = world_data 
)

summary(zib_full_2) #Even success_rate doesn't work
``` 

ZIP with time improvement
```{r}
zib_full_2 = glmmTMB(
  W8_4 ~ success_rate + day_improvement,        
  ziformula = ~ success_rate + day_improvement, 
  family = binomial(link = logit),
  data  = world_data 
)

summary(zib_full_2) 
``` 

Firth's bias reduction
```{r}
firth_simple = logistf(W8_4 ~ success_rate, data  = world_data )
summary(firth_simple) #finally a significant p-val in just using success rate

#With time
firth_time = logistf(W8_4 ~ success_rate + day_improvement, data  = world_data )
summary(firth_time)

firth_time_2 = logistf(W8_4 ~ day_improvement, data  = world_data )

summary(firth_time_2)

anova(firth_simple, firth_time)

```
Block for testing firth_simple
Actually a good fit, decent predictions, Nifski has around a 21% completion rate on average for all previous levels. world_data set updated with predictions from this.

```{r}
firth_simple = logistf(W8_4 ~ success_rate, data  = world_data )
summary(firth_simple)

# 1. Generate predictions for existing world_data 
world_data $predicted_prob = predict(firth_simple, type = "response")


# 2. DIAGNOSTIC PLOT 3: Success rate vs probability curve
ggplot(world_data , aes(x = success_rate, y = predicted_prob)) +
  geom_point(aes(color = as.factor(W8_4)), alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(title = "Success Rate vs Predicted Probability",
       x = "Success Rate (prior levels)",
       y = "Predicted Probability of World Record",
       color = "Actual WR") +
  scale_color_manual(values = c("0" = "red", "1" = "green"),
                     labels = c("No WR", "WR")) +
  theme_minimal()

# 3.goodness of fit
hoslem.test(world_data $W8_4, world_data $predicted_prob, g = 10)
cat("Test: p >0.05 indicates good fit\n")


target_row = world_data [
  world_data $W1_1 == 1 &
  world_data $W1_2 == 1 &
  world_data $W4_1 == 1 &
  world_data $W4_2 == 1 &
  world_data $W8_1 == 1 &
  world_data $W8_2 == 1 &
  world_data $W8_3 == 1,
]
target_row
world_data
```




